\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data\\using R packages}}

### 2012-01-16 package/smutans/inst/doc/smutans.Rnw
Just an update.

>>>>

\section{Prewriting about UA159 versus TW1}

\subsection{Abstract}
A bacterial transcriptome of the primary etiological agent of human dental caries,
\textit{Stretptococcus mutans}, is described using RNA-seq data. Putative
transcripts boundaries are defined. Small RNAs are predicted by computational
methods, and confirmed by RNA-seq and Southern blot experiments. We refine the
previously established importance of CcpA regulation as central metabolism and
virulence gene expression using RNA-seq data.
CcpA is known to play a direct role in carbon catabolite repression (CCR).
We describe genes that are differentially expressed in glucose-grown wild-type
(UA159) and CcpA-deficient (TW1) strains. We found that more
genes were differentially expressed in the comparison between wild-type and
CcpA-deficient strains in galactose-grown strains than in
glucose-grown strains.
We also found that more genes differentially expressed in wild-type than in
CcpA-deficient strain when comparing glucose- and galactose-grown cells. These
two findings were different from those of \cite{Abranches2008}. \fixme{We may
want to see effects of different p-value cutoff.} We confirmed that CcpA
modulates the pathogenic potential of \textit{S.\ mutans} through global control
of gene expression.
We predicted a few small RNAs based on sequence alignments and RNA secondary
structures. We confirmed them experimentally. Target genes of predicted small
RNAs were predicted. The target genes were enriched in some functional
categories. We listed highly potential small RNAs with their target genes
functional categories. 
Transcripts were predicted using RNA-seq short reads mapped on the wild-type
UA159 genome.

\subsection{Reviews of papers that cited \cite{Abranches2008}}

CCR in bacteria involves regulatory networks that activate or silence genes in
response to carbon source and availability.

\subsection{Goal: Refinement of \cite{Abranches2008}}

Comparisons of microarray and RNA-seq are not good way to go.

\subsection{Methods and Materials}

\subsubsection{Bacterial strains and growth conditions}

\subsubsection{RNA isolation}

\subsubsection{RNA sequencing}

\subsubsection{Statistical analysis for differential expressions}
We used an R package, DESeq, \cite{Anders2010} to determine differentially
expressed genes based on negative binomial model originally developed by \cite{Robinson2007}.
Steps of analyzing RNA-seq data for determining differentially expressed genes
are described in \cite{Oshlack2010}. We used a short read aligner called
\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
aligner allowed a few gaps for efficient alignment of millions of sequences of
size being 100 base pairs long, short reads that consist of RNAs and sequencing
adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
remove parts of adapter sequences from short reads before mapping on the
reference genome. Mapped short read alignments were converted into readable
formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
reference genome, forming a table of read counts of the genes. Statistical
software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
was subsequently employed to infer differentially expressed in different
biological conditions.  We also generated tracks for our recently released
Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
summarize gene expression levels based the short read alignments. This allowed
us to visualize parts of the reference genome for comparing different RNA
samples. Further analyses such as functional category enrichment test
\cite{Young2010} were also employed.

\subsubsection{small RNAs}

RNAz is used to find small RNA candidates. Sequence alignments were a
critical step for finding small RNAs in the approach. Phylogenies of streptococcus species could
be considered to choose species related with Streptococcus mutans.
Alternatively, I find sequences similar to intergenic regions, and refine the
alignments using a multiple sequence alignment tool such as MUSCLE
\cite{Edgar2004a}. Steps for
achieving this goal include: 1. Exract intergenic regions including up- and
down-stream, 2. BLAST the intergenic regions against non-redundant DB to find
similar sequences \cite{Altschul1990}, 3. Refine each alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the
alignments for scoring regions for small RNAs
\cite{Washietl2005,Gruber2010}, and 5. Find transcripts with high
score of small RNAs. The transcripts must be defined in advance.

RNAplex
\cite{Tafer2008}

A review paper about small RNA predictions and targets
\cite{Tafer2008a}

RNAplfold
\cite{Bernhart2006}

\subsubsection{Transcripts}

``-c 10 -b 25 -force_gp''

\subsubsection{Gene category associations and virulence genes}

We assigned genes to Gene Ontology (GO) categories by comparing the {\em
  Streptococcus} genes to bacterial proteins from the Uniref90 database
using BLASTP, and then assigning the same GO classification as the target
gene of the uniProt GOA database if the match had an $E$-value of
$<1.0\times10^{-5}.$  
A gene family was assigned a given
classification if any of its genes was assigned that classification.  To
test for associations with replacing gene transfers we used the
recombination intensity estimated for each gene, and for additive gene
transfers we used the number of additive transfers inferred for each
family.  To test for significance, we performed a Mann-Whitney $U$ test of
the values (recombination intensities or numbers of transfers) associated
with a given category vs.\ the values for the all other genes/families.  We
used the \cite{Benjamini1995} method to correct for multiple comparisons.

\subsection{Results}

\subsection{Discussion}

\subsection{Abbreviations}

\begin{enumerate}
\item CCR Carbon catabolite repression
\item CcpA ?
\item IPS intracellular polysaccharide
\end{enumerate}
